---
title: "TufteGraphics"
author: "B. Moran"
date: "5 October 2016"
output: 
  html_document: 
    fig_caption: yes
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, error=FALSE,fig.width=10*1.414, fig.height=10, width=10*1.414, cache = TRUE, tidy = TRUE)

#needs::needs(magrittr,grid,CarletonStats,devtools,epanetReader,fmsb,ggplot2,ggthemes,latticeExtra,MASS,PerformanceAnalytics,psych,dplyr,plyr,prettyR,plotrix,proto,RCurl,reshape,reshape2,knitr,ggExtra)
```

# Minimal Line Plot

## Base Graphics

```{r mlpbase}
x <- 1967:1977
y <- c(0.5,1.8,4.6,5.3,5.3,5.7,5.4,5,5.5,6,5)
# pdf(width=10,
#     height=6
#     )
plot(y ~ x,
     axes=F,
     xlab="",
     ylab="",
     pch=16,
     type="b"
     )
axis(1,
     at=x,
     label=x,
     tick=F,
     family="serif"
     )
axis(2,
     at=seq(1,6,1),
     label=sprintf("$%s", seq(300,400,20)),
     tick=F,
     las=2,
     family="serif"
     )
abline(h=6,
       lty=2
       )
abline(h=5,
       lty=2
       )
text(max(x),
     min(y)*2.5,
     "Per capita\nbudget expanditures\nin constant dollars",
     adj=1,
     family="serif"
     )
text(max(x),
     max(y)/1.08,
     labels="5%",
     family="serif"
     )
```

## Lattice

```{r mlplattice}
x <- 1967:1977
y <- c(0.5,1.8,4.6,5.3,5.3,5.7,5.4,5,5.5,6,5)

xyplot(y~x,
       xlab="",
       ylab="",
       pch=16,
       col=1,
       border = "transparent",
       type="o",
       abline=list(
         h = c(
           max(y),
           max(y)-1),
         lty = 2
         ),
       scales=list(
         x=list(
           at=x,
           labels=x,
           fontfamily="serif",
           cex=1
           ),
         y=list(
           at=seq(1,6,1),
           fontfamily="serif",
           cex=1,
           label=sprintf(
             "$%s",
             seq(300,400,20)
             )
           )
         ),
       par.settings = list(
         axis.line = list(
           col = "transparent"),
         dot.line=list(lwd=0)
         ),
       axis = function(side, line.col = "black", ...) {
         if(side %in% c("left","bottom")) {
           axis.default(side = side, line.col = "black", ...)
           }
         }
       )
ltext(current.panel.limits()$xlim[2]/1.1, 
      adj=1,
      fontfamily="serif", 
      current.panel.limits()$ylim[1]/1.3, 
      cex=1,
      "Per capita\nbudget expandures\nin constant dollars"
      )
ltext(current.panel.limits()$xlim[2]/1.1,
      adj=1, fontfamily="serif", 
      current.panel.limits()$ylim[1]/5.5,
      cex=1,
      "5%"
      )
```

## ggplot2

```{r mlpggplot2}
x <- 1967:1977
y <- c(0.5,1.8,4.6,5.3,5.3,5.7,5.4,5,5.5,6,5)

d <- data.frame(x, y)

ggplot(d,
       aes(x,y)
       ) + 
  geom_line() + 
  geom_point(size=3) + 
  theme_tufte(base_size = 15) + 
  theme(axis.title=element_blank()) + 
  geom_hline(yintercept = c(5,6),
             lty=2
             ) + 
  scale_y_continuous(breaks=seq(1, 6, 1),
                     label=sprintf("$%s",seq(300,400,20))
                     ) + 
  scale_x_continuous(breaks=x,
                     label=x
                     ) + 
  annotate("text", 
           x = c(1977,1977.2),
           y = c(1.5,5.5),
           adj=1,
           family="serif",
           label = c(
             "Per capita\nbudget expandures\nin constant dollars",
             "5%"
             )
           )
```

# Range-frame (or quartile-frame) scatterplot 

## Base Graphics

```{r rfsbase}
x <- mtcars$wt
y <- mtcars$mpg

plot(x,
     y,
     main="",
     axes=FALSE,
     pch=16,
     cex=0.8,
     family="serif",
     xlab="Car weight (lb/1000)",
     ylab="Miles per gallon of fuel"
     )

axis(1,
     at=summary(x),
     labels=round(summary(x),1),
     tick=F,
     family="serif"
     )

axis(2,
     at=summary(y),
     labels=round(summary(y),1),
     tick=F,
     las=2,
     family="serif"
     )
```

## fancyaxis

```{r rfsfancyaxis}
source("fancyaxis.R")
x <- mtcars$wt
y <- mtcars$mpg

plot(x, 
     y, 
     main="", 
     axes=FALSE, 
     pch=16, 
     cex=0.8,
     xlab="Car weight (lb/1000)", 
     ylab="Miles per gallon of fuel"
     )
fancyaxis(1,
          summary(x),
          digits=1
          )
fancyaxis(2, 
          summary(y), 
          digits=1
          )
```

## lattice

```{r rfslattice}
x <- mtcars$wt
y <- mtcars$mpg

xyplot(y ~ x, 
       mtcars, 
       col=1, 
       pch=16, 
       fontfamily="serif",
       xlab="Car weight (lb/1000)",
       ylab="Miles per gallon of fuel",
       par.settings = list(
         axis.line = list(col="transparent"),
         par.xlab.text=list(fontfamily="serif"),
         par.ylab.text=list(fontfamily="serif")
         ),
       scales = list(
         x=list(
           at=summary(mtcars$wt),
           labels=round(
             summary(mtcars$wt),1),
           fontfamily="serif"),
         y=list(
           at=summary(mtcars$mpg),
           labels=round(summary(mtcars$mpg),1),
           fontfamily="serif")
         ),
       axis = function(side, line.col = "black", ...) {
         if(side %in% c("left","bottom")) {
           axis.default(
             side = side,
             line.col = "black",
             ...
             )
           }
         }
       )
```


## ggplot2

```{r rfsggplot2}
ggplot(mtcars,
       aes(wt, mpg)) + 
  geom_point() + 
  geom_rangeframe() + 
  theme_tufte() +
  xlab("Car weight (lb/1000)") + 
  ylab("Miles per gallon of fuel") + 
  theme(
    axis.title.x = element_text(vjust=-0.5),
    axis.title.y = element_text(vjust=1.5)
    )
```


# Dot-dash (or rug) scatterplot
## fancyaxis

```{r ddfancyaxis}
source("fancyaxis.R")
x <- mtcars$wt
y <- mtcars$mpg

plot(x,
     y,
     main="", 
     axes=FALSE, 
     pch=16, 
     cex=0.8,
     xlab="Car weight (lb/1000)", 
     ylab="Miles per gallon of fuel", 
     xlim=c(min(x)-0.2, 
            max(x)+0.2
            ),
     ylim=c(min(y)-1.5, 
            max(y)+1.5)
     )
axis(1, tick=F)
axis(2, tick=F, las=2)
minimalrug(x, side=1, line=-0.8)
minimalrug(y, side=2, line=-0.8)
```

## lattice

```{r ddlattice}
x <- mtcars$wt
y <- mtcars$mpg

xyplot(y ~ x, 
       xlab="Car weight (lb/1000)",
       ylab="Miles per gallon of fuel",
       par.settings = list(
         axis.line = list(col="transparent")
         ),
       panel = function(x, y,...) {
         panel.xyplot(x,
                      y,
                      col=1,
                      pch=16
                      )
         panel.rug(x,
                   y,
                   col=1,
                   x.units = rep("snpc", 2),
                   y.units = rep("snpc", 2),
                   ...)
         }
       )
```


## ggplot2

```{r ddggplot2}
ggplot(mtcars, 
       aes(wt, mpg)
       ) + 
  geom_point() + 
  geom_rug() + 
  theme_tufte(ticks=F) + 
  xlab("Car weight (lb/1000)") + 
  ylab("Miles per gallon of fuel") + 
  theme(
    axis.title.x = element_text(vjust=-0.5),
    axis.title.y = element_text(vjust=1)
    )
```

# Marginal histogram scatterplot 

## fancyaxis

```{r mhsfancyaxis}
source("fancyaxis.R")
x <- faithful$waiting
y <- faithful$eruptions

plot(x, 
     y, 
     main="", 
     axes=FALSE, 
     pch=16, 
     cex=0.8,
     xlab="Time till next eruption (min)", 
     ylab="Duration (sec)", 
     xlim=c(min(x)/1.1, 
            max(x)
            ), 
     ylim=c(min(y)/1.5, 
            max(y)
            )
     )
axis(1, tick=F)
axis(2, tick=F, las=2)
axisstripchart(faithful$waiting, 1)
axisstripchart(faithful$eruptions, 2)
```

## ggplot2 with ggMarginal

```{r mhsggplot2marghist}
p <- ggplot(faithful,
            aes(waiting, eruptions)
            ) + 
  geom_point() + 
  theme_tufte(ticks=F)
ggMarginal(p, 
           type = "histogram", 
           fill="transparent"
           )
```

## As above but with marginal densities

```{r mhsggplot2margdens}
p <- ggplot(faithful, 
            aes(waiting, eruptions)
            ) + 
  geom_point() + 
  theme_tufte(ticks=F) +
  theme(
    axis.title=element_blank(),
    axis.text=element_blank()
    )
ggMarginal(p,
           type = "density"
           )
```

## As above but with boxplots

```{r mhsggplot2margboxplots}
p <- ggplot(faithful, 
            aes(waiting, eruptions)
            ) + 
  geom_point() + 
  theme_tufte(ticks=F) +
  theme(
    axis.title=element_blank(), 
    axis.text=element_blank()
    )
ggMarginal(p,
           type = "boxplot",
           size=10,
           fill="transparent"
           )
```

# Minimal Boxplots

## Base Graphics

```{r mbpbase}
x <- quakes$mag
y <- quakes$stations
boxplot(y ~ x, 
        main = "", 
        axes = FALSE, 
        xlab=" ", 
        ylab=" ",
        pars = list(
          boxcol = "transparent",
          medlty = "blank",
          medpch=16,
          whisklty = c(1, 1),
          medcex = 0.7,  
          outcex = 0, 
          staplelty = "blank")
        )
axis(1, 
     at=1:length(unique(x)), 
     label=sort(unique(x)),
     tick=F,
     family="serif"
     )
axis(2, 
     las=2, 
     tick=F, 
     family="serif"
     )
text(min(x)/3,
     max(y)/1.1,
     pos = 4, 
     family="serif",
     "Number of stations \nreporting Richter Magnitude\nof Fiji earthquakes (n=1000)"
     )
```

## chart.Boxplot

```{r mbpcB}
d <- msq[,80:84]
chart.Boxplot(d, 
              main = "", 
              xlab="average personality rating (based on n=3896)",
              ylab="", 
              element.color = "transparent", 
              as.Tufte=TRUE
              )
```

## lattice

```{r mbplattice}
x <- quakes$mag
y <- quakes$stations

bwplot(y ~ x,
       horizontal=F,
       xlab="", 
       ylab="", 
       do.out = FALSE,
       box.ratio = 0,
       scales=list(
         x=list(labels=sort(unique(x)),
                fontfamily="serif"),
         y=list(fontfamily="serif")
         ),
       par.settings = list(axis.line = list(col = "transparent"),
                           box.umbrella=list(lty=1, col= 1),
                           box.dot=list(col= 1),
                           box.rectangle = list(col=c("transparent"))
                           )
       )
ltext(#current.panel.limits()$xlim[1] + 250,
      600,
      adj=1,
      # current.panel.limits()$ylim[2] + 50,
      120,
      fontfamily="serif",
      "Number of stations \nreporting Richter Magnitude\nof Fiji earthquakes (n=1000)"
      )
```

## ggplot2

```{r mbpggplot2}
ggplot(quakes,
       aes(factor(mag),stations)
       ) + 
  theme_tufte() +
  geom_tufteboxplot(outlier.colour="transparent") + 
  theme(axis.title=element_blank()) +
  annotate("text", 
           x = 8, 
           y = 120, 
           adj=1,  
           family="serif",
           label = c("Number of stations \nreporting Richter Magnitude\nof Fiji earthquakes (n=1000)")
           )
```

# Minimal barchart

## Base Graphics

```{r mbcbase}
d <- colMeans(msq[,c(2,7,34,36,42,43,46,55,68)],na.rm = T)*10
barplot(d,
        xaxt="n",
        yaxt="n",
        ylab="", 
        border=F, 
        width=c(.35),
        space=1.8
        )
axis(1,
     at=(1:length(d))-.26,
     labels=names(d),
     tick=F,
     family="serif"
     )
axis(2,
     at=seq(1, 5, 1),
     las=2, 
     tick=F, 
     family="serif"
     )
abline(h=seq(1, 5, 1),
       col="white",
       lwd=3
       )
abline(h=0, 
       col="gray",
       lwd=2
       )
text(min(d)/2,
     max(d)/1.2,
     pos = 4, 
     family="serif",
     "Average scores\non negative emotion traits\nfrom 3896 participants\n(Watson et al., 1988)"
     )
```

## lattice

```{r mbclattice}
d <- colMeans(msq[,c(2,7,34,36,42,43,46,55,68)],na.rm = T)*10
barchart(sort(d),
         xlab="", 
         ylab="", 
         col = "grey",
         origin=1,  
         border = "transparent", 
         box.ratio=0.5, 
         panel = function(x,y,...) {
           panel.barchart(x,y,...)
           panel.abline(v=seq(1,6,1), 
                        col="white", lwd=3)
           },
         par.settings = list(axis.line = list(col = "transparent")))
ltext(current.panel.limits()$xlim[2]-50,
      adj=1,  
      current.panel.limits()$ylim[1]-100,
      "Average scores\non negative emotion traits\nfrom 3896 participants\n(Watson et al., 1988)"
      )
```

## ggplot2

```{r mbcggplot2}
d <- melt(colMeans(msq[,c(2,7,34,36,42,43,46,55,68)],na.rm = T)*10)
d$trait <- rownames(d)
ggplot(d,
       aes(x=trait, y=value)
       ) + 
  theme_tufte(base_size=14, ticks=F) +
  geom_bar(width=0.25, 
           fill="gray", 
           stat = "identity") +  
  theme(axis.title=element_blank()) +
  scale_y_continuous(breaks=seq(1, 5, 1)) + 
  geom_hline(yintercept=seq(1, 5, 1), 
             col="white", 
             lwd=1
             ) +
  annotate("text", 
           x = 3.5, 
           y = 5, 
           adj=1,  
           family="serif", 
label = c("Average scores\non negative emotion traits
          from 3896 participants\n(Watson et al., 1988)")
)
```

# Slopegraph 

## Base Graphics

```{r sgbase}
source("slopegraph.r")
d <- read.csv("Data/cancer_survival_rates.csv")
df <- reshape(d,timevar = "year",idvar = c("group"),direction = "wide")
rownames(df) <- df$group; df <- round(df[,-1],0)
slopegraph(df, col.line='gray', binval = 2, labels=c("5 years", "10 years", "15 years", "20 years"), col.xaxt="transparent", main="")
```

## Base & Bump Chart

```{r sgbasebump}
source("slopegraph.r")
d <- read.csv("Data/cancer_survival_rates.csv", header = TRUE)

df <- stretch_df(d, 
                 "group", 
                 "value"
                 )
bumpchart(df[,3:6],
          rank=FALSE,
          col="gray50",
          top.labels=c(5,10,15,20),
          labels=df[,1],
          main="Years cancer survival by group", 
          mar=c(2,12,5,12)
          )
```

## ggplot2

```{r sgggplot2}
library(ggplot2)
library(ggthemes)
library(devtools)
library(RCurl)
library(plyr)
source("slopegraph2.R")
d <- read.csv("Data/cancer_survival_rates.csv")
df <- transform(d, x=factor(year, levels=c(5,10,15,20), labels=c("5 years","10 years","15 years","20 years")), y=round(value))
df <- build_slopegraph(df = df, x="year", y="value",group= "group", method="tufte", min.space=0.04)
plot_slopegraph(df) + 
  labs(title="Estimates of % survival rates") +
  theme_tufte(base_size=16, ticks=F) + 
  theme(axis.title=element_blank())
```

# Sparklines

## Base Graphics

```{r slbase}
dd <- read.csv("Data/us_nr_of_crimes_1960_2014.csv")
d <- dd[,c(2:11)]
#pdf("sparklines_base.pdf", height=10, width=6)
par(mfrow=c(ncol(d),1), 
    mar=c(1,0,0,8), 
    oma=c(4,1,4,4)
    )
for (i in 1:ncol(d)){
  plot(d[,i],
       lwd=0.5,
       axes=F,
       ylab="",
       xlab="",
       main="",
       type="l",
       new=F
       )
  axis(4,
       at=d[nrow(d),i],
       labels=round(d[nrow(d),i]),
       tick=F, 
       las=1, 
       line=-1.5, 
       family="serif", 
       cex.axis=1.2
       )
  axis(4, 
       at=d[nrow(d),i], 
       labels=names(d[i]),
       tick=F,
       line=1.5, 
       family="serif",
       cex.axis=1.4,
       las=1
       )
  text(which.max(d[,i]), 
       max(d[,i]),
       labels=round(max(d[,i]),0), 
       family="serif",
       cex=1.2,
       adj=c(0.5,3)
       )
  text(which.min(d[,i]), 
       min(d[,i]), 
       labels=round(min(d[,i]),0), 
       family="serif",
       cex=1.2,
       adj=c(0.5,-2.5)
       )
  ymin <- min(d[,i])
  tmin <- which.min(d[,i])
  ymax <- max(d[,i])
  tmax <- which.max(d[,i])
  points(x=c(tmin,tmax),
         y=c(ymin,ymax),
         pch=19,
         col=c("red","blue"),
         cex=1
         )
  rect(0, 
       summary(d[,i])[2],
       nrow(d), 
       summary(d[,i])[4],
       border=0, 
       col = rgb(190, 190, 190, alpha=90, maxColorValue=255)
  )}
axis(1,
     at=1:nrow(dd), 
     labels=dd$Year, 
     pos=c(-5),
     tick=F, 
     family="serif", 
     cex.axis=1.4
     )
#dev.off()
```

## plotSparklineTable

```{r slplSpLTbl}
dd <- read.csv("Data/us_nr_of_crimes_1960_2014.csv")
d <- melt(dd[,c(2:11)])
#pdf("sparklines_base_epanetReader.pdf", height=6, width=10)
plotSparklineTable(d, row.var = 'variable', col.vars = 'value')
#dev.off()
```

## lattice

```{r sllattice}
dd <- read.csv("Data/us_nr_of_crimes_1960_2014.csv")
d <- melt(dd, id="Year")
names(d)[1] <- "time"
#pdf("sparklines_lattice.pdf", height=10, width=8)
xyplot(value~time | variable, 
       d, 
       xlab="",
       ylab="",
       strip=F, 
       lwd=0.7, 
       col=1,
       type="l", 
       layout=c(1,length(unique(d$variable))), 
       between = list(y = 1),
       scales=list(
         y=list(at=NULL, relation="free"), 
         x=list(fontfamily="serif")
         ), 
       par.settings = list(axis.line = list(col = "transparent"),
                           layout.widths=list(right.padding=20, 
                                              left.padding=-5)
                           ),
       panel = function(x, y, ...) {
         panel.xyplot(x, y, ...)
         pushViewport(
           viewport(xscale=current.viewport()$xscale-5,
                    yscale=current.viewport()$yscale, 
                    clip="off")
           )
         panel.text(x=tail(x,n=1), 
                    y=tail(y,n=1), 
                    labels=levels(d$variable)[panel.number()], 
                    fontfamily="serif", 
                    pos=4
                    )
         popViewport()
         panel.text(x=x[which.max(y)], 
                    y=max(y), 
                    labels=round(max(y),0),
                    cex=0.8,
                    fontfamily="serif",
                    adj=c(0.5,2.5)
                    )
         panel.text(x=x[which.min(y)], 
                    y=min(y), 
                    labels=round(min(y),0), 
                    cex=0.8,
                    fontfamily="serif",
                    adj=c(0.5,-1.5)
                    )
         panel.text(x=tail(x,n=1), 
                    y=tail(y,n=1), 
                    labels=round(tail(y,n=1),0),
                    cex=0.8,
                    fontfamily="serif",
                    pos=4
                    )
         panel.points(x[which.max(y)],
                      max(y),  
                      pch=16,
                      cex=1
                      )
         panel.points(x[which.min(y)],
                      min(y), 
                      pch=16,
                      cex=1, 
                      col="red"
                      )
         }
       )
#dev.off()
```

## ggplot2

```{r slggplot2}
dd <- read.csv("Data/us_nr_of_crimes_1960_2014.csv")
d <- melt(dd, id="Year")
names(d) <- c("Year","Crime.Type","Crime.Rate")
d$Crime.Rate <- round(d$Crime.Rate,0)
mins <- group_by(d, Crime.Type) %>% slice(which.min(Crime.Rate))
maxs <- group_by(d, Crime.Type) %>% slice(which.max(Crime.Rate))
ends <- group_by(d, Crime.Type) %>% filter(Year == max(Year))

quarts <- d %>% 
  dplyr::group_by(Crime.Type) %>%
  dplyr::summarise(
    quart1 = quantile(Crime.Rate, 0.25),
    quart2 = quantile(Crime.Rate, 0.75)
    ) %>%
  right_join(d)

# pdf("sparklines_ggplot.pdf", height=10, width=8)

ggplot(d,aes(x=Year, y=Crime.Rate)) + 
  facet_grid(Crime.Type ~ ., 
             scales = "free_y"
             ) + 
  geom_ribbon(data = quarts, 
              aes(ymin = quart1, 
                  max = quart2
                  ), 
              fill = 'grey90'
              ) +
  geom_line(size=0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins,
            aes(label = Crime.Rate),
            vjust = -1
            ) +
  geom_text(data = maxs,
            aes(label = Crime.Rate),
            vjust = 2.5
            ) +
  geom_text(data = ends,
            aes(label = Crime.Rate), 
            hjust = 0, 
            nudge_x = 1
            ) +
  geom_text(data = ends,
            aes(label = Crime.Type),
            hjust = 0,
            nudge_x = 5
            ) +
  expand_limits(x = max(d$Year) + (0.25 * (max(d$Year) - min(d$Year)))) +
  scale_x_continuous(breaks = seq(1960, 2010, 10)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 15, 
              base_family = "Helvetica"
              ) +
  theme(axis.title=element_blank(), 
        axis.text.y = element_blank(), 
        axis.ticks = element_blank(), 
        strip.text = element_blank()
        )
#dev.off()
```

# Stem-and-leaf display

## Base Graphics (in Console)

```{r saldbase}
stem(faithful$eruptions)
```

## CarletonStats (in Console)

```{r saldCarletonStats}
stemPlot(birthwt$bwt, 
         birthwt$smoke, 
         varname="infant birth weight (in grams)",
         grpvarname="whether mother smoked during pregnancy (1) or not (0)"
         )
```

## Base Graphics & fmsb

```{r saldbasefmsb}
gstem(faithful$eruptions)
```